---
title: "Project_PracticalMachineLearning"
author: "Maja Kocic"
date: "12 november 2016"
output: html_document
---
## Goal
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.


##Prediction Assignment

##Background

Using devices such as JawboneUp, NikeFuelBand, and Fitbitit is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.

In this project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

###Preparing the data and R packages

```{r echo=FALSE}
library(ggplot2)
library(caret)
library(randomForest)
#install.packages("gbm")
library(gbm)
#install.packages("plyr")
setwd("C:/Users/Miroslav/Desktop/CourseraMachineLearning")
#load("Project.file.ML")
```

###Getting Data
```{r echo=FALSE, comment = ""}
# URL of the training and testing data
train.url ="https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test.url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
# file names
train.name = "./pml-training.csv"
test.name = "./pml-testing.csv"
# if directory does not exist, create new
if (!file.exists("C:/Users/Miroslav/Desktop/CourseraMachineLearning")) {
  dir.create("./CourseraMachineLearning")
}
# if files does not exist, download the files
if (!file.exists(train.name)) {
  download.file(train.url, destfile=train.name, method="curl")
}
if (!file.exists(test.name)) {
  download.file(test.url, destfile=test.name, method="curl")
}
# load the CSV files as data.frame 
# load data
training <- read.csv("./pml-training.csv", row.names = 1)
testing <- read.csv("./pml-testing.csv", row.names = 1)
# remove near zero covariates
nsv <- nearZeroVar(training, saveMetrics = T)
training <- training[, !nsv$nzv]
# remove variables with more than 80% missing values
nav <- sapply(colnames(training), function(x) if(sum(is.na(training[, x])) > 0.8*nrow(training)){return(T)}else{return(F)})
training <- training[, !nav]
# calculate correlations
cor <- abs(sapply(colnames(training[, -ncol(training)]), function(x) cor(as.numeric(training[, x]), as.numeric(training$classe), method = "spearman")))
#dim(training)
#dim(testing)
#names(training)
```
The raw training data has 19622 rows of observations and 158 features (predictors). Column X is unusable row number. While the testing data has 20 rows and the same 158 features. There is one column of target outcome named classe.
###Load data
Remove near zero covariates and those with more than 80% missing values since these variables will not provide much power for prediction.
Calculate correlations between each remaining feature to the response, classe. Use spearman rank based correlation because classe is a factor.
Plot the two features that have highest correlation with classe and color with classe to see if we can separate response based on these features.

```{r,fig.width = 6, fig.height = 5}
# plot predictors 
#summary(cor)
plot(training[, names(which.max(cor))], training[, names(which.max(cor[-which.max(cor)]))], col = training$classe, pch = 19, cex = 0.1, xlab = names(which.max(cor)), ylab = names(which.max(cor[-which.max(cor)])))
```


The training set has r nrow(training) samples and r ncol(training) - 1 potential predictors after filtering.

There doesn't seem to be any strong predictors that correlates with classe well, so linear regression model is probably not suitable in this case. Boosting and random forests algorithms may generate more robust predictions for our data.

###Boosting model

Fit model with boosting algorithm and 10-fold cross validation to predict classe with all other predictors.
Plot accuracy of this model on the scale [0.9, 1].


```{r echo=FALSE, comment = ""}
set.seed(123)
boostFit <- train(classe ~ ., method = "gbm", data = training, verbose = F, trControl = trainControl(method = "cv", number = 10))
```

```{r,fig.width = 6, fig.height = 5}

boostFit
plot(boostFit, ylim = c(0.9, 1))
```

The boosting algorithm generated a good model with accuracy = 0.997.

###Random forests model

Fit model with random forests algorithm and 10-fold cross validation to predict classe with all other predictors.
Plot accuracy of the model on the same scale as boosting model.

```{r, echo=FALSE}
set.seed(123)
rfFit <- train(classe ~ ., method = "rf", data = training, importance = T, trControl = trainControl(method = "cv", number = 10))
```
```{r,fig.width = 6, fig.height = 5}
rfFit
plot(rfFit, ylim = c(0.9, 1))
imp <- varImp(rfFit)$importance
imp$max <- apply(imp, 1, max)
imp <- imp[order(imp$max, decreasing = T), ]
```

The random forests algorithm generated a very accurate model with accuracy close to 1. Compared to boosting model, this model generally has better performance in terms of accuracy as we see from the plots.

###Final model and prediction

Comparing model accuracy of the two models generated, random forests and boosting, random forests model has overall better accuracy. So, I'll use this model for prediction.
The final random forests model contains 500 trees with 40 variables tried at each split. The five most important predictors in this model are r rownames(imp)[1:5].
Estimated out of sample error rate for the random forests model is 0.04% as reported by the final model.
Predict the test set and output results for automatic grader.

```{r,echo=FALSE}
# final model
rfFit$finalModel
# prediction
(prediction <- as.character(predict(rfFit, testing)))
# write prediction files
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("./prediction/problem_id_", i, ".txt")
    write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, col.names = FALSE)
  }
}
pml_write_files(prediction)
```
